# Data wrangling


A major component of data analysis is data wrangling. This is the process of cleaning and transforming raw data into a usable format. This is a critical step in the data analysis process, as the quality of the data will directly impact the quality of the analysis. In this section, we will cover some common data wrangling tasks in R, including cleaning data, transforming data, and handling missing values.

In general this will be the most time consuming period of your analysis, cleaning the names of the variables, deleting variables, changing names of levels/factors, finding missing values, etc. A major issue will always be the missing values, and how to deal with them in addition to resolving the issues of the data such as outliers and non sensical data. In almost all project data wrangling will incolve more than 80% of your time and that data anlysis usually less than 20% of your time.

In this chapter we will cover the following topics: **Put in the same order as the code in the text**

- Cleaning column names with **janitor** and **clean_names**
- Changing the names to variables with **rename**
- Deleting variables with **select**
- Changing names of levels/factors with **fct_recode**
- Finding missing values with **is.na** and **summary**
- Subsetting data set with **filter**
- Select variables based on conditions with **select**
- Filter rows based on conditions with **filter**
- Arrange rows based on a variable with **arrange**
- Calculate new variables based on existing variables with **mutate**
- Exclude NA values with **complete.cases** and **na.omit**
- Group data by a variable with **group_by**
- Summarize data by a variable with **summarize**
- Join data frames with **left_join**, **right_join**, **inner_join**, and **full_join**
- Reshape data with **pivot_longer** and **pivot_wider**
- Export data with as .csv file with **write_csv** or a Excel file with **write_xlsx**


## Cleaning column names

### Cleaning the names of the variables with **janitor**

The easiest part is likely to clean the names of the variables (column names). This can be done with the `janitor` package. This package has a function called `clean_names` that will organize the names of the variables in a data frame with specific charatceristics. This function will convert the names to lowercase, remove special characters, and replace spaces with underscores. This will make the names of the variables more consistent and easier to work with. Sometime older functions in R will not work with variables that have spaces or special characters, such as (é, ü, &, %, ect).

Let us start with a dirty data frame with the following problems:

 - the names of the variables are in uppercase
 - the names of the variables have spaces
 - the names of the variables have special characters
 - the names of the variables are too long
 - the names of the variables are not consistent
 - the names of the variables are not informative


```{r, wrangling_1, message=FALSE, warning=FALSE, }
library(tibble)

Dirty_data=tribble(~"Name", ~"Age ^2", ~"Height (m)", ~"% of value", ~"Date of Birth", ~"Place of Birth",
                   "Alice", 25, 1.70, .70, "1997-01-01", "New York City Manhathan",
                   "Bob", 30, 1.80, .80, "1992-01-01", "Los Angeles",
                   "Charlie", 35, NA, .90, "1987-01-01", NA)
 
Dirty_data                  
```


### Now let us clean the names of the variables

 Note the following changes
 
  - all are in lowercase
  - the spaces are replaced by underscores
  - the special characters are removed and even given new names
  - the names are more informative
  - the names are more consistent


```{r, wrangling_2, warning=FALSE, message=FALSE}
library(janitor)

Clean_df_clean=clean_names(Dirty_data)

Clean_df= Clean_df_clean

Clean_df # observe the changes in the names of the variables
```


## Changing the names to variables 

To change the names of the variables in a data frame, you can use the `rename` function from the `dplyr` package. This function allows you to specify the new names of the variables using the `new_name = old_name` syntax. For example, to change the name of the `name` variable to `first_name` in a data frame called `df`, you would run:

Changing the names of the variables is a common data wrangling task that can help make your data more readable and easier to work with. By using the `rename` function, you can quickly and easily change the names of the variables in a data frame to better reflect the data they contain.


We will replace the name of the variable `name` to `full_name` in the data frame `Clean_df`.

```{r}
library(dplyr)

Clean_df <- Clean_df %>% rename(first_name = name)

Clean_df # observe the change in the name of the variable


```






## Deleting variables

We can delete variables from a data frame using the `select` function from the `dplyr` package. This function allows you to select a subset of variables from a data frame by specifying the variables you want to keep. To delete a variable from a data frame, you can simply exclude it from the list of variables you want to keep. For example, to delete the `age_2` variable from a data frame called `df`, you would run:

```{r}
library(dplyr)

Clean_df <- Clean_df %>% select(-age_2)

Clean_df # observe the change in the data frame, now the column is removed
```


## Changing names of levels/factors

We can change the names of the levels or factors in variable using the `fct_recode` function from the `forcats` package. Let us recode the name of the cities. 



```{r}
library(forcats)

Clean_df$place_of_birth <- fct_recode(Clean_df$place_of_birth,
                                      "NY" = "New York City Manhathan",
                                      "LA" = "Los Angeles")

Clean_df # observe the change in the data frame, now the names of the cities are recoded
```



## Finding missing values

One of the challenges is detecting missing value in large data sets. The `is.na` function can be used to detect missing values in a data frame. This function returns a logical vector that indicates whether each element in the data frame is missing or not. For example, to find missing values in a data frame called `df`, you would run:

Note that when there is missing data, the function will return a logical vector with `TRUE` for missing values and `FALSE` for non-missing values.

```{r}
is.na(Clean_df) # this will return a logical vector indicating whether each element in the data frame is missing or not
```

The following code is acceptible for small data set, but for large data sets it is better to use the `summary` function. This function will return a summary of the data frame, including the number of missing values in each variable. For example, to get a summary of the missing values in a data frame called `df`, you would run:

Note here that the place of birth for Charlie is missing, it has an NA value in addition to the height_m of an individual.
```{r}
summary(Clean_df) # this will return a summary of the data frame, including the number of missing values in each variable
```

## Subsetting data with missing values

Subset all rows with missing values in the variables. This will result in a rwo that will include only Charlie. 

Subset all rows with missing values in the variables. This will result in a rwo that will include only Charlie. NOTE the the command is the **!** symbol before the `complete.cases` function. This will return all rows with missing values in the data frame.

```{r}

Clean_df[!complete.cases(Clean_df), ] # this will return all rows with missing values in the data frame


```

Note if you want subset all rows with NO missing values in the variables. This will result in a two rows that will include only Alice and Bob. 


```{r}
Clean_df[complete.cases(Clean_df), ] # this will return all rows with NO missing values in the data frame
```



## Exclude NA values from a specific variable with **drop_na**

This function will exclude missing values from a specific variable in a data frame. For example, to exclude missing values from the `place_of_birth` variable in a data frame called `Clean_df`.  Note it removes the row with the missing value in the place of birth variable.



```{r}

library(tidyverse)
Clean_df |> drop_na(place_of_birth) # this will return all rows with missing values in the place of birth variable
```



## Subsetting data set

The function to subset a data frame is `filter`. This function is used to extract rows from a data frame that meet a certain condition. The syntax for the `filter` function is as follows:


```{r, wrangling_3, warning=FALSE, message=FALSE}
library(dplyr)

Clean_df_clean %>% filter(age_2 > 29) # this will return all rows where the age is greater than 29




```

## Titanic data set

```{r}
library(readr)
Titanic <- read_csv("Data/ Titanic.csv") # The data is in the "DATA" folder

Titanic = clean_names(Titanic) # this will clean the names of the variables in the data frame
names(Titanic) # this will return the names of the variables in the data frame
```



Let us now filter all rows with females, and that are in the first class and age less than 20. 

From 1,309 passengers we 16 females with the above characteristics.

```{r}
Titanic %>% 
  filter(sex=="female" & pclass==1 & age<20) # this will return

# The following code those the same as the above code, but it is more readable for beginners

Titanic %>% 
  filter(sex=="female" & age<20) |> arrange(age)


Titanic %>% 
  filter(sex == "female") |> 
  filter(pclass == 1) |>
  filter(age < 20) # this will return
```

### Arrange rows based on a variable

To arrange the rows of a data frame based on a variable, you can use the `arrange` function from the `dplyr` package. This function allows you to sort the rows of a data frame based on the values of a variable. The syntax for the `arrange` function is as follows:

Note now the data frame is sorted by the age variable. 


```{r}
library(dplyr)

Clean_df_clean %>% arrange(age_2) # this will return all rows sorted by the age variable
```


### Mutate: Calculate new variables based on existing variables

To calculate new variables based on existing variables in a data frame, you can use the `mutate` function from the `dplyr` package. This function allows you to create new variables by performing calculations on existing variables. The syntax for the `mutate` function is as follows:

Let us calculate the BMI of the individuals in the data frame. 

- First of all let us add the missing value for the height of Charlie. 
- The add a column of the weight in pounds of all individuals.


```{r}
library(dplyr)

Clean_df_clean <- Clean_df_clean %>% mutate(height_m = ifelse(name == "Charlie", 1.75, height_m)) # this will return all rows with a new variable called height_m

Clean_df_clean <- Clean_df_clean %>% mutate(weight_kg = ifelse(name == "Alice", 60, ifelse(name == "Bob", 70, 80))) # this will return all rows with a new variable called weight_kg


```

Now use mutate to calculate the BMI of the individuals in the data frame. 

Note we use the same notattion as mathematical notation, the `^` symbol is used to raise a number to a power. the variable is given the name of `bmi`

```{r}
Clean_df_clean <- Clean_df_clean %>% mutate(bmi = weight_kg / height_m^2) # this will return all rows with a new variable called bmi

Clean_df_clean # this will return the data frame with the new variable bmi


```

### Group data by a variable

To group a data frame by a variable, you can use the `group_by` function from the `dplyr` package. This function allows you to group the rows of a data frame based on the values of a variable. The syntax for the `group_by` function is as follows:

Let us group the data frame by the `sex` variable and calculate the mean age of the individuals in each group.



```{r}
library(dplyr)

Titanic %>% group_by(sex) |>  # this will return all rows grouped
  summarise(mean_age = mean(age, na.rm = TRUE)) # this will return the mean age of the individuals in the data frame

```

### Group by multiple variables. 

Now we will summarize the by group by the `sex` and `pclass` variables and calculate the mean age and minimum and maximum age of the individuals in each subgroup in the data frame. 




```{r}
Titanic %>% 
  group_by(sex, survived, pclass) |>
  summarise(mean_age = mean(age, na.rm = TRUE),
            min_age = min(age, na.rm=TRUE),
            max_age = max(age, na.rm=TRUE)) # this will return the mean age of the individuals in the data frame
```

