# Test of normality



Add and activate packages.
NOTE THE NEW FUNCTION to install all packages ONLY IF Needed.....pacman::p_load(car, pastecs, psych, ggplot2)

```{r packages, include=FALSE}
#----Install and Load Packages-----
#install.packages("car", dependencies = TRUE)
#install.packages("ggplot2")
#install.packages("pastecs")
#install.packages("psych")
if (!require("pacman")) install.packages("pacman")
pacman::p_load(car, pastecs, psych, ggplot2, nortest)


library(car) # Companion to applied regression
library(ggplot2)  # Data Visualization
library(pastecs) # Space time Ecological Series
library(psych)   # Procedures for Psychological, psychometric and Personality Research
library(nortest) # Package for normality test
#library(modeest) # package to calculate the mode of a dataset
```



```{r histogram}

library(readr)
DownloadFestival <- read_csv("Data/DownloadFestival.csv")

head(DownloadFestival)

dlf <- DownloadFestival
```

### METADATA
The data is from a festival in UK. The data is about the hygiene score of the participants. The hygiene score is a score from 0 to 4.

The scale is as follows:

 - 0 = You smell like a rotting corpse
 - 4 = You smell like of sweet roses



Multiple variables

```{r multiple variables}
#Two alternative ways to describe multiple variables.
psych::describe(cbind(dlf$day1, dlf$day2, dlf$day3))
psych::describe(dlf[,c("day1", "day2", "day3")])

```

## Shapiro-Wilk normality test


Test of normality, Shapiro Wilks Test is a test of normality in frequentist statistics. It tests the null hypothesis that the data was drawn from a normal distribution. The test is based on the correlation between the data and the corresponding normal scores. The test is considered to be the most powerful test for normality, particularly for small sample sizes. However it is sensitive to the presence of outliers. and tends to be conservative when the sample size is large. in other words it tends to reject the null model when sample size are large and the data is sufficient normal.



```{r test of normality}

library(pastecs)
stat.desc(dlf$day3, basic = FALSE, norm = TRUE) # "norm=TRUE"" is to calculate the Shapiro Wilk Test
```

```{r description by variable}
stat.desc(cbind(dlf$day1, dlf$day2, dlf$day3), basic = FALSE, norm = TRUE)
```

## The shapiro-Francia normality test

The Shapiro-Francia test is known to perform well, see also the comments by Royston (1993). The expected ordered quantiles from the standard normal distribution are approximated by qnorm(ppoints(x, a = 3/8)), being slightly different from the approximation qnorm(ppoints(x, a = 1/2)) used for the normal quantile-quantile plot by qqnorm for sample sizes greater than 10.

Royston, P. (1993): A pocket-calculator algorithm for the Shapiro-Francia test for non-normality: an application to medicine. Statistics in Medicine, 12, 181–184.

Thode Jr., H.C. (2002): Testing for Normality. Marcel Dekker, New York.

```{r}
sf <- sf.test(dlf$day3)
sf 

```



## Anderson-Darling normality test

The Anderson-Darling test is a modification of the Kolmogorov-Smirnov test that gives more weight to the tails of the distribution. It is a more powerful test than the Kolmogorov-Smirnov test, but it is also more sensitive to departures from normality in the center of the distribution.

Stephens, M.A. (1986): Tests based on EDF statistics. In: D'Agostino, R.B. and Stephens, M.A., eds.: Goodness-of-Fit Techniques. Marcel Dekker, New York.

Thode Jr., H.C. (2002): Testing for Normality. Marcel Dekker, New York.


```{r}
ad.test(dlf$day1)
```


## Cramer-von Mises test of normality

The Cramer-von Mises test is a modification of the Anderson-Darling test that gives more weight to the center of the distribution. It is a more powerful test than the Anderson-Darling test, but it is also more sensitive to departures from normality in the tails of the distribution.

Stephens, M.A. (1986): Tests based on EDF statistics. In: D'Agostino, R.B. and Stephens, M.A., eds.: Goodness-of-Fit Techniques. Marcel Dekker, New York.

Thode Jr., H.C. (2002): Testing for Normality. Marcel Dekker, New York.

```{r}
cvm.test(dlf$day1)
```

## Lilliefors test of normality

The Lilliefors (Kolomorov-Smirnov) test is the most famous EDF omnibus test for normality. Compared to the Anderson-Darling test and the Cramer-von Mises test it is known to perform worse. Although the test statistic obtained from lillie.test(x) is the same as that obtained from ks.test(x, "pnorm", mean(x), sd(x)), it is not correct to use the p-value from the latter for the composite hypothesis of normality (mean and variance unknown), since the distribution of the test statistic is different when the parameters are estimated.

Dallal, G.E. and Wilkinson, L. (1986): An analytic approximation to the distribution of Lilliefors' test for normality. The American Statistician, 40, 294–296.

Stephens, M.A. (1974): EDF statistics for goodness of fit and some comparisons. Journal of the American Statistical Association, 69, 730–737.

Thode Jr., H.C. (2002): Testing for Normality. Marcel Dekker, New York.

```{r}
lillie.test(dlf$day1)
```




## Visualizing the data

### qq-plot

Add a straight line on the qqplot
```{r qqplot and line}

# This function is to add a straight line through the qqplot
qqplot.data <- function (vec) # argument: vector of numbers
{
  # following four lines from base R's qqline()
  y <- quantile(vec[!is.na(vec)], c(0.25, 0.75))
  x <- qnorm(c(0.25, 0.75))
  slope <- diff(y)/diff(x)
  int <- y[1L] - slope * x[1L]
  
  d <- data.frame(resids = vec)
  
  ggplot(d, aes(sample = resids)) + 
    stat_qq() + 
    geom_abline(slope = slope, intercept = int, color="red")
  
}

qqplot.data(dlf$day3)


ggsave("qqplot.png")
```


### Histogram

```{r}


dlf=subset(DownloadFestival)
head(dlf)
tail(dlf)

hist.day1 <- ggplot(dlf, aes(day1)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") 

hist.day1+ 
  labs(x="Hygiene score on day 1", y = "Density")

hist.day1

ggsave("histogram_festival_hygiene.pdf") # Can be either be a device function (e.g. png()), or one of "eps", "ps", "tex" (pictex), "pdf", "jpeg", "tiff", "png", "bmp", "svg" or "wmf" (windows only)

```


### Density plot

The Normal Distribution

https://en.wikipedia.org/wiki/Normal_distribution


1. visualization if the observed distribution follows a theoretical normal distribution
2. Test to determine if the observed distribution follows a theoretical distribution


 $$P(x)=\frac{1}{{\sigma\sqrt{ 2\pi}}}{e}^{\frac{{(x-µ)}^{2}}{{2\sigma}^{2}}}$$
 




```{r adding the normal curve}
# Ahora añadir la linea de distribución normal

hist.day1 + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(dlf$day1,na.rm = TRUE), 
                sd = sd(dlf$day1 , na.rm = TRUE)), 
                colour = "blue", size = 1)

shapiro.test(dlf$day1)  # don't use with more than 40 a 200 data point
length(dlf$day1)
```












Visualize the distribution of the data "Histogram"







#Remove the outlier from the day1 hygiene score
dlf$day1 <- ifelse(dlf$day1 > 5, NA, dlf$day1)
df$Column = ifelse(df$column_to_be evaluated, replace_with_NA, otherwise_leave_as_before)


```{r remove outlier}
dlf$day1 <- ifelse(dlf$day1 > 5, NA, dlf$day1)


# Note here that we use ..density..  # CUAL ES LA DIFERENCIA entre densidad y frecuencia?  

hist.day1 <- ggplot(dlf, aes(day1)) + 
  theme(legend.position = "none") + 
  geom_histogram(aes(y=..density..), colour="black", fill="white") + 
  labs(x="Hygiene score on day 1", y = "Density")

hist.day1

```











```{r quantifying normality}

dlf=DownloadFestival
#Quantifying normality with numbers
library(psych)		#load the "psych" library, if you haven't already, for the describe() function.

#Using the describe() function for a single variable.
psych::describe(dlf$day2)

```
Que es la varianza?
The variance

$$s^{ 2 }=\frac { \sum _{ i=1 }^{ n }{ (x_{ i }-\bar { x } ) } ^{ 2 } }{ n-1 } $$

Que es la desviación estandard

The standard deviation of the mean

$$s=\sqrt { s^{ 2 } } $$











Dicover the Mode. La moda en R

```{r mode}
# the mode
library(modeest)

mfv(dlf$day1, method="mfv")
mfv(dlf$day2, method="mfv")
mfv(dlf$day3, method="mfv")



```




Test of normality, Shapiro Wilks Test

```{r test of normality SW}

library(pastecs)
stat.desc(dlf$day3, basic = FALSE, norm = TRUE) # "norm=TRUE"" is to calculate the Shapiro Wilk Test
```



```{r rounding}
round(stat.desc(dlf[, c("day1", "day2", "day3")], basic = FALSE, norm = TRUE), digits = 3)

```

# Levene's Test: testing for assumption of normality among groups

```{r levene test}
#Levene's test for comparison of variances of exam scores in the two universities.
library(ggplot2)
library(car)
library(readr)
RExam <- read_csv("Data/RExam.csv")

rexam=RExam
head(rexam)

ggplot(rexam, aes(numeracy, fill=as.factor(uni)))+
  geom_histogram()+
  facet_wrap(~uni)



 leveneTest(rexam$lectures, rexam$uni)
leveneTest(rexam$lectures, rexam$uni, center = mean)

leveneTest(rexam$computer, rexam$uni)
leveneTest(rexam$numeracy, rexam$uni)




```




```{r transformation}

x=c(0,2,3,4,5,6,100,1000)
x
log(x+1)
sqrt(x)

head(log(dlf$day1))

head(dlf)
dlf$lgday1=log(dlf$day1)
head(dlf)
```




