# Multinomial logit analysis

A multinomial logit model is a type of regression analysis used to model nominal outcome variables. In this model, the dependent variable is a nominal variable with more than two levels. The model is used to estimate the probability of each level of the nominal variable. The model is an extension of the binary logit model, which is used to model binary outcome variables. The multinomial logit model is used in a wide range of fields, including economics, political science, and marketing.

The model is based on the assumption that the log-odds of each level of the nominal variable are a linear function of the predictor variables. The model is estimated using maximum likelihood estimation, and the coefficients of the model are interpreted as the change in the log-odds of each level of the nominal variable associated with a one-unit change in the predictor variable.

In this tutorial, we will walk through an example of how to estimate a multinomial logit model in R using the `nnet` package. We will use the `iris` dataset, which is a built-in dataset in R that contains measurements of iris flowers. The outcome variable in the dataset is the species of the iris flower, which has three levels: setosa, versicolor, and virginica. We will estimate a multinomial logit model to predict the species of the iris flower based on the measurements of the flower.

### Load libraries

```{r}
library(tidyverse)
library(nnet)


```


### Data

The data used in this analysis is the `iris` dataset, which is a built-in dataset in R. The dataset contains 150 observations of iris flowers, with four predictor variables (sepal length, sepal width, petal length, and petal width) and one outcome variable (species of iris flower). The outcome variable has three levels: setosa, versicolor, and virginica.

```{r}
data(iris)

head(iris)


```

### Model


### Assumptions

### Post estimation analysis




