# Poisson Regression

```{r, PR_1, echo=FALSE, message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,  huxtable,  SuppDists, univariateML, invgamma)


#detach("package::lubridate", unload=TRUE)
library(tidyverse) # this is a package that includes dplyr, ggplot2, tidyr, readr, purrr, tibble, stringr, forcats
library(huxtable) # this is a package to create tables
library(SuppDists) # this is a package to create distributions
library(univariateML) # this is a package to create distributions
library(invgamma) # this is a package to create distributions

# library(wakefield) # this is a package to create random data: (DO NOT activate it as there multiple conflicts wit other packages), Use is this alternative wakefield::r_sample_binary(50, x = 0:1, prob =c(0.3, 0.7), name = "Binary")

```

A poisson regression is used when the dependent variable is a count variable, that is, the number of events that occur in a fixed period of time. The Poisson distribution is used as a probability distribution and the **log** function is used as a link function. The model is called **Poisson Regression**.

$$\log\left(\lambda_i\right)=\beta_0+\beta_1X_i$$
where the distribution is a Poisson with $y_i\sim Poisson\left(\lambda_i\right)$.

Here let use the data from the number of accidents in a city and the number of cars in the city.

 - Brooklyn_Bridge:  The number cyclist on the Brooklyn Bridge on each day.
 - Precipitation = The precipitation in inches on that day
 
 
```{r, PR_2, warning=FALSE, message=FALSE}
library(readr)
NY_CITY_CYCLIST <- read_csv("Data/NY_CITY_CYCLIST.csv")

NY=NY_CITY_CYCLIST

head(NY)
```
 
 
## Visualize the data

- The data is not normally distributed.
- The data is a count variable.
- The regression line is not linear and is count (number of cylcist)
- The 95% confidence interval is shown in the graph.


```{r, PR_3, warning=FALSE, message=FALSE}
ggplot(NY, aes(x=Precipitation, y=Brooklyn_Bridge))+
  geom_jitter(width=0.01, height = 0.1)+ # point represents a count of number of cyclist for a specific day
  geom_smooth(method="glm", method.args=list(family="poisson"), se=T)+ 
    xlab("Precipitation in inches")+
  ylab("Number of cyclist")+
  scale_x_continuous(breaks = seq(0, 2, by = 0.1))
```

## Creating a Poisson model


```{r, PR_4, warning=FALSE, message=FALSE}
NY_model=glm(Brooklyn_Bridge ~ Precipitation, data=NY, family=poisson)
summary(NY_model)
```

## Overdispersion in a Poisson model

The Poisson model assumes that the mean and variance are equal.  If the variance is greater than the mean, then the model is overdispersed.  In this case, we will use the function **dispersiontest** from the package **AER** to test for overdispersion. The function will return the dispersion parameter and the p-value. If the p-value is less than 0.05, then there is overdispersion.



Testing for overdispersion of Poisson model.  In this case the model is highly obverdisperded. The dispersion is 418 and the p-value is p < 0.001.


```{r, PR_5, warning=FALSE, message=FALSE}

library(AER)
dispersiontest(NY_model)



```




****

## Negative binomial regression

A negative binomial regression is used when the dependent variable is a count variable and that the data suggests overdipersion.  Overdispersion is when the variance is greater than the mean. The Poisson distribution is used as a probability distribution and the **log** function is used as a link function. The model is called **Negative Binomial Regression**.

$$\log\left(\lambda_i\right)=\beta_0+\beta_1X_i$$
where the distribution is a negative binomial with $y_i\sim NegBinomial\left(\lambda_i\right)$.

We will start by calculating the mean and vaiance of the above data set and detemrine if there is overdisperion

The function required **dispersiontest** is available in the package **AER**. The function will return the dispersion parameter and the p-value. If the p-value is less than 0.05, then there is overdispersion.


```{r, PR_6, warning=FALSE, message=FALSE}
library(AER)
model=glm(Brooklyn_Bridge ~ Precipitation, data=NY, family=poisson)


```



Note that the alpha is very large and the p-value is less than 0.05, which indicates that there is overdispersion. We will now create the negative binomial model. We will use the MASS library and the function **glm.nb** to create the model. The function is used to create a negative binomial model. The function is used in the same way as the glm function, but the family is set to negative binomial.

## Visualization of the negative binomial model

The graph shows the fitted line of the negative binomial model. The graph also shows the 95% confidence interval of the fitted line. The graph also shows the data points and the jitter of the data points. The graph also shows the x-axis and y-axis labels.

```{r, PR_7, warning=FALSE, message=FALSE}
library(MASS)

ggplot(NY, aes(x=Precipitation, y=Brooklyn_Bridge))+
  geom_jitter(width=0.01, height = 0.1)+
  geom_smooth(method="glm.nb", se=T)+ 
    xlab("Precipitation in inches")+
  ylab("Number of cyclist")+
  scale_x_continuous(breaks = seq(0, 2, by = 0.1))
```

## Creating the negative binomial model 



```{r, PR_8, warning=FALSE, message=FALSE}
NY_model_nb=MASS::glm.nb(Brooklyn_Bridge ~ Precipitation, data=NY)

summary(NY_model_nb) # note that theta is a the measure of dispersion
```

We interpret the results as follows

- The precipitation has a negative effect on the number of cyclist (negative sign). With precipitation (exp(coef) = -0.53) which states that there is 47% decrease in the number of cyclist for **each inch of precipitation**. 

We can reconvert the coefficient to a positive value by taking the exponential of the coefficient. 

```{r, PR_9, warning=FALSE, message=FALSE}
exp(coef(NY_model_nb))
```


In this case to evaluate overdispersion in the negative binomial model we use the function **odTest** from the package **pscl**. The function will return the dispersion parameter and the p-value. If the p-value is less than 0.05, then there is overdispersion.



```{r, PR_10, warning=FALSE, message=FALSE}
library(pscl)

odTest(NY_model_nb)
```

****
### Comparing dispersion models

Comparing the dispersion of a Poisson and a Negative binomial model. The approach is to use a likelyhood ratio test.  Note the fitted log-likelihood of the NB model (-1677.1) is much large/better than the Poisson model (-41692.7). The p-value is less than 0.05, thus the negative binomial model is better than the Poisson model.

```{r, PR_11, warning=FALSE, message=FALSE}
    # Calculate the likelihood ratio statistic
    likelihood_ratio <- 2 * (logLik(NY_model_nb) - logLik(NY_model))
    
logLik(NY_model_nb) # the log likelihood of the negative binomial model
logLik(NY_model) # the log likelihood of the Poisson model


    # Degrees of freedom (df) is the difference in the number of parameters
    df <- length(coef(NY_model_nb)) - length(coef(NY_model))
    
    # Perform the chi-squared test
    p_value <- 1 - pchisq(likelihood_ratio, df)
    
    # Print the p-value
    print(p_value)
```

### Alternative approach to compare the Poisson and the negative binomial model.

You can also use the function **lrtest** from the package **lmtest** to perform the likelihood ratio test. The function will return the likelihood ratio test statistic and the p-value. If the p-value is less than 0.05, then the negative binomial model is better than the Poisson model.

```{r, PR_12, warning=FALSE, message=FALSE}
library(lmtest)
lrtest(NY_model_nb, NY_model)
```


## Overlaying the Poisson and the Negative binomial models

In this case the Poisson model is shown in red and the negative binomial model is shown in blue. The graph shows the fitted line of the Poisson model and the negative binomial model. The graph also shows the 95% confidence interval of the fitted line. The graph also shows the data points and the jitter of the data points. The graph also shows the x-axis and y-axis labels. There is a statistical diference between the models, however the fitted lines are very similar and consequently in this case the differences may not be important. 
 



```{r, PR_13, warning=FALSE, message=FALSE}
ggplot(NY, aes(x=Precipitation, y=Brooklyn_Bridge))+
  geom_jitter(width=0.01, height = 0.1)+
  geom_smooth(method="glm", method.args=list(family="poisson"), se=T, color="red")+ 
  geom_smooth(method="glm.nb", se=T, color="blue")+ 
    xlab("Precipitation in inches")+
  ylab("Number of cyclist")+
  scale_x_continuous(breaks = seq(0, 2, by = 0.1))
```

